---
title: "Annual trends in California fisheries from 1992-2014"
author: "Sophia Wassermann"
date: "3/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data source

These data were downloaded from the [California Department of Natural Resources, Characterization of Human Uses and the Socioenonomic Dimensions, California North Coast MPA Baseline Study, 1992 to 2014][1].


## Overall timeseries
Looking at the timeseries for the number of fishers, total revenue, and pounds landed from 1992-2014, there is a general declining trend in fishers and pounds landed. Revenue is highly variable over the timeseries, with a large increase in 2012 and 2014.

```{r, include=FALSE}
# Code for a summary of the timeseries data for all of CA, pulled from all_ca_summary.R
library(reshape2)
library(ggplot2)

ca_summary <- read.csv("Data/ca_summary_92-14.csv")

# Overall trends across timeseries --------------------------------------------
summary_long <- melt(ca_summary, id="year")  # convert to long form for faceted graph
summary_long$year <- as.numeric(summary_long$year)

# Create list to use as year labels in graph
years <- c("92", "93", "94", "95", "96", "97", "98", "99", "00", "01", "02", 
           "03", "04", "05", "06", "07", "08", "09", "10", "11", "12", "13", "14")

summary_plot <- ggplot(summary_long, aes(x = year, y = value)) + 
  theme_bw() +
  geom_line(size = 1.5) +  
  ylab(" ") + 
  scale_x_continuous(labels=years, breaks=ca_summary$year) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  facet_wrap(~variable, scale="free", ncol = 1)
```

```{r show_figure, fig.width = 4.5, fig.height = 5.5, fig.align = "center", echo=FALSE}
summary_plot
```


[1]: https://data.cnra.ca.gov/dataset/human-uses-and-socioeconomic-dimensions-ca-north-coast-mpa-baseline-study-1992-2014
